Spatial modelling of the electoral two-party preferred vote in Australia. A study of federal elections between 2001 and 2016 via the \texttt{eechidna} \texttt{R} package.

### 2019-03-26 _book/electoral-modelling-draft.tex
cache

  \title{Spatial modelling of the electoral two-party preferred vote in Australia. A study of federal elections between 2001 and 2016 via the \texttt{eechidna} \texttt{R} package.}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Jeremy Forbes, Di Cook \& Rob Hyndman}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
      \predate{\centering\large\emph}
  \postdate{\par}
    \date{2019-03-24}

\usepackage{booktabs}
\usepackage{amsthm}
\makeatletter
\def\thm@space@setup{%
  \thm@preskip=8pt plus 2pt minus 4pt
  \thm@postskip=\thm@preskip
}
\makeatother

@@

\emph{insert table here}

@@

>>>>

  \title{Spatial modelling of the electoral two-party preferred vote in Australia. A study of federal elections between 2001 and 2016 via the \texttt{eechidna} \texttt{R} package.}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Jeremy Forbes, Di Cook \& Rob Hyndman}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
      \predate{\centering\large\emph}
  \postdate{\par}
    \date{2019-03-26}

\usepackage{booktabs}
\usepackage{amsthm}
\makeatletter
\def\thm@space@setup{%
  \thm@preskip=8pt plus 2pt minus 4pt
  \thm@postskip=\thm@preskip
}
\makeatother
\usepackage{longtable}

@@

\begin{table}[!htbp] \centering 
  \caption{Estimated model for each of the six elections.} 
  \label{} 
\scriptsize 
\begin{tabular}{@{\extracolsep{1pt}}lcccccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{6}{c}{\textit{Dependent variable:}} \\ 
\cline{2-7} 
\\[-1.8ex] & \multicolumn{6}{c}{Two-party preferred vote in favor of the Liberal party} \\ 
 & 2001 & 2004 & 2007 & 2010 & 2013 & 2016 \\ 
\\[-1.8ex] & (1) & (2) & (3) & (4) & (5) & (6)\\ 
\hline \\[-1.8ex] 
 $\rho$ & 0.44$^{***}$ & 0.15$^{*}$ & $-$0.04 & 0.15 & 0.29 & 0.48$^{***}$ \\ 
  & (0.15) & (0.17) & (0.19) & (0.17) & (0.16) & (0.16) \\ 
  & & & & & & \\ 
 AusCitizen & $-$2.77 & $-$2.83 & 0.67 & 1.00 & $-$0.87 & $-$1.53 \\ 
  & (2.51) & (2.63) & (2.57) & (2.93) & (2.90) & (2.95) \\ 
  & & & & & & \\ 
 Born\_Asia & 2.63 & 0.74 & $-$1.37 & $-$7.58$^{***}$ & $-$2.56 & $-$2.13 \\ 
  & (3.03) & (2.55) & (1.96) & (2.68) & (2.58) & (2.20) \\ 
  & & & & & & \\ 
 Born\_MidEast & 1.92 & 3.22$^{*}$ & 2.92$^{*}$ & 1.79 & 1.70 & $-$0.82 \\ 
  & (1.53) & (1.78) & (1.51) & (1.54) & (1.54) & (1.56) \\ 
  & & & & & & \\ 
 Born\_SE\_Europe & 1.02 & 0.04 & $-$0.20 & $-$1.72 & $-$1.25 & $-$1.63 \\ 
  & (2.34) & (2.05) & (1.17) & (1.34) & (1.15) & (0.99) \\ 
  & & & & & & \\ 
 Born\_UK & $-$0.49 & $-$0.87 & 0.29 & $-$0.29 & $-$0.35 & $-$1.37 \\ 
  & (1.04) & (0.90) & (0.80) & (1.05) & (1.00) & (0.98) \\ 
  & & & & & & \\ 
 BornElsewhere & $-$1.08 & $-$0.51 & 5.01 & 9.78$^{**}$ & 4.04 & 1.74 \\ 
  & (3.40) & (3.67) & (3.59) & (4.47) & (4.46) & (4.52) \\ 
  & & & & & & \\ 
 Buddhism & 1.50 & 2.84$^{*}$ & 0.90 & 2.41$^{*}$ & 1.09 & 1.51 \\ 
  & (1.94) & (1.61) & (1.25) & (1.40) & (1.25) & (1.24) \\ 
  & & & & & & \\ 
 Christianity & 0.53 & 3.44 & $-$1.93 & $-$0.60 & 3.08 & 5.17$^{**}$ \\ 
  & (4.58) & (4.19) & (3.45) & (3.44) & (2.79) & (2.52) \\ 
  & & & & & & \\ 
 CurrentlyStudying & $-$1.14 & 1.69 & 4.03$^{***}$ & 3.28$^{***}$ & 1.73 & $-$0.09 \\ 
  & (1.01) & (1.09) & (1.12) & (1.21) & (1.17) & (1.13) \\ 
  & & & & & & \\ 
 DeFacto & $-$6.42$^{***}$ & $-$5.04$^{**}$ & $-$6.55$^{***}$ & $-$8.60$^{***}$ & $-$9.73$^{***}$ & $-$11.02$^{***}$ \\ 
  & (1.69) & (2.00) & (1.90) & (2.57) & (2.62) & (2.55) \\ 
  & & & & & & \\ 
 DiffAddress & 3.86$^{***}$ & 4.63$^{***}$ & 3.40$^{***}$ & 4.72$^{***}$ & 5.02$^{***}$ & 6.93$^{***}$ \\ 
  & (0.95) & (1.03) & (0.88) & (1.60) & (1.71) & (1.50) \\ 
  & & & & & & \\ 
 Distributive & 1.67 & 2.75$^{**}$ & 2.25$^{**}$ & 2.88$^{**}$ & 2.76$^{**}$ & 1.65 \\ 
  & (1.08) & (1.13) & (1.02) & (1.30) & (1.25) & (1.17) \\ 
  & & & & & & \\ 
 Education & 0.62 & 1.94 & $-$0.77 & $-$1.82 & $-$0.90 & $-$3.04 \\ 
  & (2.23) & (2.86) & (2.89) & (3.35) & (3.04) & (2.99) \\ 
  & & & & & & \\ 
 Extractive & 4.69$^{***}$ & 5.31$^{***}$ & 5.81$^{***}$ & 6.97$^{***}$ & 6.09$^{***}$ & 6.81$^{***}$ \\ 
  & (1.36) & (1.23) & (1.16) & (1.48) & (1.42) & (1.31) \\ 
  & & & & & & \\ 
 FamHouseSize & $-$2.94 & $-$5.11$^{*}$ & $-$9.67$^{***}$ & $-$8.22$^{**}$ & $-$6.16$^{*}$ & $-$4.35 \\ 
  & (2.27) & (2.78) & (2.66) & (3.28) & (3.15) & (2.86) \\ 
  & & & & & & \\ 
 Incomes & 2.99$^{*}$ & 2.95 & 4.36$^{*}$ & 2.73 & 4.19 & 8.46$^{***}$ \\ 
  & (1.76) & (2.60) & (2.56) & (3.06) & (2.68) & (2.57) \\ 
  & & & & & & \\ 
 Indigenous & 3.97$^{***}$ & 5.62$^{***}$ & 6.81$^{***}$ & 5.90$^{***}$ & 6.12$^{***}$ & 4.47$^{***}$ \\ 
  & (1.26) & (1.35) & (1.17) & (1.49) & (1.49) & (1.38) \\ 
  & & & & & & \\ 
 Islam & 0.15 & 0.61 & $-$1.21 & $-$1.04 & 1.20 & 2.64 \\ 
  & (2.03) & (2.00) & (1.80) & (1.94) & (1.71) & (1.72) \\ 
  & & & & & & \\ 
 Judaism & 1.02 & 1.74 & 0.01 & 0.46 & 1.60$^{*}$ & 1.96$^{***}$ \\ 
  & (1.38) & (1.26) & (1.00) & (1.05) & (0.85) & (0.71) \\ 
  & & & & & & \\ 
 ManagerAdminClericalSales & 2.10$^{***}$ & 3.28$^{***}$ & 5.21$^{***}$ & 4.32$^{***}$ & 3.99$^{***}$ & 4.62$^{***}$ \\ 
  & (0.70) & (0.85) & (0.78) & (1.00) & (0.99) & (1.03) \\ 
  & & & & & & \\ 
 Married & 2.27 & 5.50$^{**}$ & 5.06$^{*}$ & 5.07$^{*}$ & 1.46 & $-$1.39 \\ 
  & (2.26) & (2.79) & (2.59) & (2.91) & (2.71) & (2.64) \\ 
  & & & & & & \\ 
 MedianAge & 1.54 & 3.73$^{**}$ & 1.86 & 2.08 & 2.28 & 2.94$^{*}$ \\ 
  & (1.30) & (1.55) & (1.65) & (2.08) & (1.88) & (1.63) \\ 
  & & & & & & \\ 
 NoReligion & 1.06 & 3.43 & $-$0.31 & 0.25 & 3.53 & 5.76$^{**}$ \\ 
  & (3.85) & (3.45) & (2.90) & (2.98) & (2.58) & (2.59) \\ 
  & & & & & & \\ 
 OneParent\_House & $-$1.00 & 1.37 & 1.66 & 0.22 & 0.11 & $-$0.85 \\ 
  & (1.34) & (1.49) & (1.38) & (1.61) & (1.46) & (1.39) \\ 
  & & & & & & \\ 
 OtherLanguageHome & $-$10.40$^{**}$ & $-$6.76 & $-$6.02 & $-$4.37 & $-$4.69 & 1.27 \\ 
  & (4.97) & (5.61) & (4.44) & (5.70) & (5.50) & (4.95) \\ 
  & & & & & & \\ 
 PropertyOwned & $-$0.76 & $-$0.92 & 0.78 & $-$1.17 & 0.68 & 1.76 \\ 
  & (1.19) & (1.21) & (1.09) & (1.49) & (1.43) & (1.38) \\ 
  & & & & & & \\ 
 RentLoanPrice & $-$0.31 & $-$1.54 & $-$0.54 & 2.12 & 1.01 & $-$0.70 \\ 
  & (1.50) & (1.68) & (1.48) & (1.96) & (1.96) & (2.02) \\ 
  & & & & & & \\ 
 SocialServ & 2.42$^{*}$ & 1.89 & 1.77 & 1.91 & 0.94 & 2.90$^{***}$ \\ 
  & (1.27) & (1.27) & (1.14) & (1.35) & (1.22) & (1.10) \\ 
  & & & & & & \\ 
 Transformative & 2.99$^{**}$ & 5.74$^{***}$ & 5.20$^{***}$ & 3.28$^{*}$ & 3.74$^{**}$ & 4.75$^{***}$ \\ 
  & (1.41) & (1.59) & (1.47) & (1.74) & (1.57) & (1.41) \\ 
  & & & & & & \\ 
 Unemployment & $-$3.00$^{**}$ & $-$5.09$^{***}$ & $-$2.88$^{**}$ & $-$2.42 & $-$1.22 & 0.36 \\ 
  & (1.28) & (1.51) & (1.41) & (1.73) & (1.54) & (1.42) \\ 
  & & & & & & \\ 
 Constant & 50.78$^{***}$ & 52.63$^{***}$ & 47.32$^{***}$ & 49.92$^{***}$ & 53.48$^{***}$ & 50.45$^{***}$ \\ 
  & (0.67) & (0.45) & (0.35) & (0.51) & (0.56) & (0.74) \\ 
  & & & & & & \\ 
\hline \\[-1.8ex] 
Observations & 150 & 150 & 150 & 150 & 150 & 150 \\ 
Log Likelihood & $-$398.33 & $-$403.34 & $-$398.73 & $-$418.08 & $-$405.41 & $-$399.52 \\ 
Akaike Inf. Crit. & 860.67 & 870.68 & 861.46 & 900.16 & 874.83 & 863.03 \\ 
Bayesian Inf. Crit. & 949.60 & 959.61 & 950.40 & 989.10 & 963.76 & 951.96 \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Note:}  & \multicolumn{6}{l}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
 & \multicolumn{6}{l}{Estimated coefficients for variable named in column one shown for} \\ 
 & \multicolumn{6}{l}{election year indicated by column heading, with estimated standard} \\ 
 & \multicolumn{6}{l}{deviation for each coefficient shown below in parenthesis. Overall} \\ 
 & \multicolumn{6}{l}{summary measures for each regression equation are provided in the} \\ 
 & \multicolumn{6}{l}{bottom panel.} \\ 
\end{tabular} 
\end{table}

@@

\hypertarget{spatial-autoregressive-parameter}{%
\section{Spatial autoregressive parameter}\label{spatial-autoregressive-parameter}}

\begin{center}\includegraphics{electoral-modelling-draft_files/figure-latex/rhovis-1} \end{center}

The spatial autoregressive coefficient \(\rho\) is positive and significant in only the 2001 and 2016 elections, meaning that in these elections, an electorate's political climate was affected by the attitudes of it's neighbours. Conversely, in the other four elections, it appears that after accounting for electoral socio-demographics, there is no spatial effect, meaning electorates effectively acted independently.

\hypertarget{influential-socio-demographics}{%
\section{Influential socio-demographics}\label{influential-socio-demographics}}

To investigate the socio-demographics that have a strong effect on the two-party preferred vote, partial residual plots are used. These show the direction, size and significance of an estimated effect, as the slope of the prediction line matches the estimated coefficient, and the shaded region represents a 95\% confidence band. If a horizontal line can be drawn through the confidence band, then the effect is insignificant. Plots for each election are faceted to compare the effects over time.

\emph{It is important here to note the ecological fallacy - insights are being drawn at the electorate level, and cannot be inferred for another disaggregate level (e.g.~individual voters).}

\hypertarget{income-and-unemployment}{%
\subsection{Income and unemployment}\label{income-and-unemployment}}

Typically the Labor party campaigns on more progressive policies, which often include tax reform that adversely affects higher income earners, and more generous social assistance programs. Although topical in the lead up to an election, incomes only significantly affected electoral two-party preference in 2016 (see Figure \ref{fig:plotincomes}), with higher income areas associated with support for the Liberal party. In the prior elections, the estimated effect is still positive, albeit insignificant. Unemployment on the other hand, was influential in 2001, 2004 and 2007 elections(Figure \ref{fig:plotunemploy}), with electorates with higher unemployment favouring Labor. This estimated effect has become weaker over time, with unemployment having virtually no effect in 2016.

{\centering \includegraphics[width=0.6\linewidth]{electoral-modelling-draft_files/figure-latex/plotincomes-1}

\caption{Effect of Incomes}\label{fig:plotincomes}
\end{figure}

{\centering \includegraphics[width=0.6\linewidth]{electoral-modelling-draft_files/figure-latex/plotunemploy-1}

\caption{Effect of Unemployment}\label{fig:plotunemploy}
\end{figure}

\hypertarget{industry-and-type-of-work}{%
\subsection{Industry and type of work}\label{industry-and-type-of-work}}

Electorates with higher proportions of workers in mining, gas, water, agriculture, waste and electricity (grouped as \texttt{Extractive} industries) are consistently linked with higher support for the Liberal party, with the mangitude of this effect slightly increasing over the years (see Figure \ref{fig:plotextractive}). This is unsurprising, as the Liberal party has close ties with these traditional energy industries, and typically present policies to reduce taxation on energy production. Furthermore, electorates with more workers in construction or manufacturing industries (\texttt{Transformative}), as well as wholesale trade, retail trade, transport, post or warehousing (\texttt{Distributive}), are also linked with the Liberal party. However, these effects are not as strong, and \texttt{Transformative} is marginally insignificant in the 2010 election, whereas \texttt{Distributive} was not a key socio-demographic in 2001 or 2016. These effects are shown in Figures \ref{fig:plottransformative} and \ref{fig:plotdistributive}.

{\centering \includegraphics[width=0.6\linewidth]{electoral-modelling-draft_files/figure-latex/plotextractive-1}

\caption{Effect of Extractive}\label{fig:plotextractive}
\end{figure}

{\centering \includegraphics[width=0.6\linewidth]{electoral-modelling-draft_files/figure-latex/plottransformative-1}

\caption{Effect of Transformative}\label{fig:plottransformative}
\end{figure}
\begin{figure}[h]

{\centering \includegraphics[width=0.6\linewidth]{electoral-modelling-draft_files/figure-latex/plotdistributive-1}

\caption{Effect of Distributive}\label{fig:plotdistributive}
\end{figure}

Similarly, the proportions of workers in managerial, administrative, clerical and sales roles (\texttt{ManagerAdminClericalSales}) is also a significant predictor of two-party preference across all six elections, with higher proportions supporting the Liberal party.

{\centering \includegraphics[width=0.6\linewidth]{electoral-modelling-draft_files/figure-latex/plotmanager-1}

\caption{Effect of ManagerAdminClericalSales}\label{fig:plotmanager}
\end{figure}

\hypertarget{household-mobility}{%
\subsection{Household mobility}\label{household-mobility}}

{\centering \includegraphics[width=0.6\linewidth]{electoral-modelling-draft_files/figure-latex/plotdiffaddress-1}

\caption{Effect of DiffAddress}\label{fig:plotdiffaddress}
\end{figure}

Electorates with a higher proportion of people that have recently moved house (meaning in last five years) are more likely to support the Liberal party, with this effect becoming stronger since the 2010 election. Having controlled for characteristics of house ownership and rental prices (via the factors \texttt{PropertyOwned} and \texttt{RentLoan} respectively), this is effect is somewhat surprising. \emph{Add more here?}

\hypertarget{indigenous-population}{%
\subsection{Indigenous population}\label{indigenous-population}}

{\centering \includegraphics[width=0.6\linewidth]{electoral-modelling-draft_files/figure-latex/plotindigenous-1}

\caption{Effect of Indigenous population}\label{fig:plotindigenous}
\end{figure}

The indigenous population of an electorate is found to be a significant predictor in every election, with higher indigenous populations supporting the Liberal party. There are a few outliers with Lingiari (NT) being the most extreme. However, Lingiari itself has been a strong Labor supporter, and this effect appears to be driven by the other electorates with high Indigenous populations (Kalgoorlie (WA), Leichhardt (NSW), Durack (WA), Parkes(NSW) and Kennedy (QLD)), which have almost always elected a Liberal candidate.

\hypertarget{relationships-and-families}{%
\subsection{Relationships and families}\label{relationships-and-families}}

In each of the six elections, higher proportions people in de facto marriages are tied to support for the Labor party, whereas the effect of marriages themselves are insignificant. Larger family and household size (via the \texttt{FamHouseSize} factor) is also linked with Labor support, although only being significant in the 2007 and 2010 elections (see Figure \ref{fig:plotfamhousesize}).

\emph{Potentially remove plots}

\hypertarget{residuals}{%
\section{Residuals}\label{residuals}}

Largest residuals

Residuals by state

\begin{center}\includegraphics{electoral-modelling-draft_files/figure-latex/unnamed-chunk-2-1} \end{center}
